---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# UK2GTFS

The goal of UK2GTFS is to convert train, tram, bus, and metro timetable data from the unfriendly and difficult to use formats used in the UK to the easy to use [GTFS](https://developers.google.com/transit/gtfs/) format. The main purpose of developing the package is to support using [OpenTripPlanner](https://github.com/ITSLeeds/opentripplanner) in the UK.

## Introduction

The UK has two main sources of public transport timetable data [**traveline**](https://www.travelinedata.org.uk/) publishes data on buses and light rail and [**ATOC**](http://data.atoc.org/rail-industry-data) pubishes data on heavy rail. Each uses a data fromat that is unique to that organisation and both formats are old and difficult to use.  Many countires have adopeted the [GTFS](https://developers.google.com/transit/gtfs/) format which was developed by Google as a simple standard for publishing timetable data. Due to the wide-spread adoption of GTFS many useful tools have been develope that accept GTFS files as inputs. This package is intended to make the conversion of UK data to the GTFS format simple and easy.

## Capabilities - why we need another package

There are a number of pre-existing options for converting data to GTFS. This package aims to go beyond basic conversion by providing a range or additonal fucntionality:

1. Data conversion
    + Conversion of TransXchange to GTFS
    + Conversion of ATOC CIF files to GTFS 
1. Data cleaning, the raw data often contains clear errors, the package does not blindly convert but also corrects some known errors
    + Improved locations of tiplocs
    + Correction of mispelt organisations
    + Removal of on-demand bus services (GTFS does not support services that are on-demand)
1. Data polishing
    + Support of journeys past midnight
    + Routing of train journeys along tracks rather than straight lines between stops


## Installation

Install the package with **devtools** as follows:

```{r install, eval=FALSE}
install.packages("devtools") # If you do not already have the devtools package
devtools::install_github("ITSleeds/UK2GTFS")
```

## Package Status

This package a work in progress and comes with no guarantees. As of September 2019, it can convert most train and bus timetables to GTFS but occasionally fails on specific files. 
