---
title: "ATOC - Train Timetables"
author: "Malcolm Morgan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ATOC - Train Timetables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette outlines how to produce GTFS timetables for the heavy rail network using ATOC CIF files.

## Getting Data

Data for train timetables can be downloaded from the [Rail Data Portal](https://opendata.nationalrail.co.uk/).

```{r download, eval=FALSE}
library(UK2GTFS)
nrdp_timetable("myfolder/timetable.zip")
```

Note that you need to register a free account with the Rail Data Portal. Any you need to set the `NRDP_username` and `NRDP_password` values in in your `.Renviron` file. You can use `usethis::edit_r_environ()` to easily access your `.Renviron` file. Remember to restart R after updating the file.

You can also use the `nrdp_fares()` and `nrdp_routing()` functions to download additional data. 

## Converting ATOC Data to GTFS

To do a simple conversion to GTFS, load the package, and use the `atoc2gtfs` function. The function requires the following arguments.

* `path_in` is the file path to the ATOC timetable file, this comes a zipped folder with a name like `ttis627.zip`
* `ncores` If > 1 then multicore processing is used to speed up the processing. Always leave one spare core for your OS.

A simple example is shown below:

```{r atoc_example, eval=FALSE}

path_in <- "myfolder/timetable.zip"
gtfs <- atoc2gtfs(path_in = path_in,
          ncores = 3)
```

Once the conversion has been completed, you can save the GTFS file.

```{r atoc_example2, eval=FALSE}
gtfs_write(gtfs, 
           folder = "C:/GTFS",
           name = "gtfs_rail")

```

This will save the GTFS file to `C:/GTFS/gtfs_rail.zip`

## Alternative Data Sources

CIF Files are available from Network Rail, the formatting is slightly different and so a separate function `nr2gtfs()` is provided.

## Advanced features

The `atoc2gtfs()` function gives some customization options to how the GTFS file is built. It is also just a wrapper around other functions in the package. This section will explain the main steps.

### Importing ATOC Data

After unzipping the folder, the reading of the ATOC data is done with the following functions:


* `importMCA()` the main file with the timetable data, 
* `importFLF()` optional file used if `transfers = TRUE`
* `importMSN()` optional file used if `locations = "file"`

Other files are currently ignored.

### TIPLOC Locations

The timetable data refers to locations such as stations, junctions, etc. using a TIPLOC id. The locations of these are stored in the `.msn` file, but are not always accurate. Therefore the package comes with a `tiplocs` object which includes improved [location data for TIPLOCS](https://github.com/itsleeds/UK2GTFS-data/blob/main/data/raw/tiplocs.csv).

The additional locations and other additional data are managed by the [UK2GTFS-data](https://github.com/itsleeds/UK2GTFS-data) repo. Contributions of new or improved locations are welcome via pull request.

If you wanted to build using the locations included in the ATOC data, then you can use `locations = "file"` in `atoc2gtfs()` or manually build the stops using `station2stops()`.


### Agency 

The data required to create the GTFS `agency.txt` file is not present in the ATOC CIF files. Therefore the package contains an [example file](https://github.com/itsleeds/UK2GTFS-data/blob/main/data/raw/atoc_agency.csv) instead. Contributions to this file are welcome via pull request.
