---
title: "ATOC - Train Timetables"
author: "Malcolm Morgan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ATOC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Getting Data

### ATOC Data - for trains

Data for trains can be downloaded from http://data.atoc.org/how-to you will need to register for a free account and login.



## Using the package

### Converting ATOC Data to GTFS

Load the package, and use the `atoc2gtfs` function. The fucntnon requires the following arguments.

* `path_in` is the file path to the ATOC timetable file, this comes a ziped folder with a name like `ttis123.zip`
* `path_out` is the path to a folder where the results will be saved
* `name` is the file name of the result by default `gtfs.zip`
* `silent` if FALSE extra information is displayed duing processing
* `ncores` If > 1 then mulicore processing is used to speed up the processing
* `locations` station locations can be `tiplocs` object provided with the package (default) or taken from the ATOC `"file"` (known to have errors) or file path to CSV file.
* `agency` agency details which are not included in the network rail files, default used example data provided by the package

```{r atoc_example, eval=FALSE}
library(UK2GTFS)
#path_in = "C:/Users/Malcolm/OneDrive - University of Leeds/UK Trains Timetables/ATOC Data/2019-08-31/ttis381.zip"
#path_out = "C:/Users/Malcolm/OneDrive - University of Leeds/UK Trains Timetables/ATOC Data/2019-08-31/"
path_in = "E:/OneDrive - University of Leeds/UK Trains Timetables/ATOC Data/2019-08-31/ttis381.zip"
path_out = "E:/OneDrive - University of Leeds/UK Trains Timetables/ATOC Data/2019-08-31/"
atoc2gtfs(path_in = path_in, 
          path_out = path_out,
          name = "gtfs4",
          silent = FALSE, 
          ncores = 6)

```





## Making a GTFS file from the ATOC Data



```{r setup, eval=FALSE}
path_in = "E:/OneDrive - University of Leeds/UK Trains Timetables/ATOC Data/2019-08-31/ttis381.zip"
path_out = "E:/OneDrive - University of Leeds/UK Trains Timetables/ATOC Data/2019-08-31/"
ncores = 5
silent = FALSE
source("../R/atoc.R")
source("../R/atoc_export.R")
source("../R/atoc_import.R")
source("../R/atoc_main.R")
atoc2gtfs(path_in = path_in, path_out = path_out, silent = silent, ncores = ncores)
```

```{r setup2, eval=FALSE}
path_in = "~/data/ttis508.zip"
path_out = "~/data"
ncores = 5
silent = FALSE
name = "gtfs"
library(UK2GTFS)
load("../data/rail_heavy.rda")
source("../R/atoc.R")
source("../R/atoc_export.R")
source("../R/atoc_import.R")
source("../R/atoc_main.R")
#timetables <- atoc2gtfs(path_in = path_in, path_out = path_out, silent = silent, ncores = ncores)
#saveRDS(timetables,"ATOC_timetables.Rds")
gtfs <- readRDS("ATOC_timetables.Rds")
shapes <- trips2shapes(gtfs = gtfs,  ncores = ncores)


```



## Making Shapes 

```{r, eval=FALSE}
trips = read.csv("D:/Users/earmmor/OneDrive - University of Leeds/Routing/gtfs/trips.txt",, stringsAsFactors = F)
routes = read.csv("D:/Users/earmmor/OneDrive - University of Leeds/Routing/gtfs/routes.txt", stringsAsFactors = F)
stop_times = read.csv("D:/Users/earmmor/OneDrive - University of Leeds/Routing/gtfs/stop_times.txt", stringsAsFactors = F)
#stops = read.csv("D:/Users/earmmor/OneDrive - University of Leeds/Routing/gtfs/stops.txt")
stops = sf::st_read("D:/Users/earmmor/OneDrive - University of Leeds/Routing/osm/stops_fixed.shp")
```


