---
title: "Heavy Rail CIF to GTFS - Train Timetables"
author: "Malcolm Morgan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Heavy Rail CIF to GTFS - Train Timetables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explains how to convert Great Britain's **heavy rail** timetables
into [GTFS](https://gtfs.org/), where the source data comes from, and why there
are two different conversion functions for what looks like the same format.

## Background: the CIF format

Great Britain's national rail timetable is published in the **Common Interface
File** (CIF) format - a fixed-width text format dating back to the mainframe era.
Each line is a record whose meaning is given by its first two characters (for
example `BS` = basic schedule, `LO` = origin location, `LI` = intermediate
location, `LT` = terminating location). It is compact and complete but
thoroughly unfriendly to modern tools: there are no column headers, positions
are counted in fixed character offsets, and locations are given as **TIPLOC**
codes (Timing Point Locations) rather than coordinates.

A full timetable download is actually a bundle of related files:

| Extension | Contents | Used by UK2GTFS |
|-----------|----------|-----------------|
| `.MCA` | The main timetable (schedules, calls, times) | **Yes** (required) |
| `.MSN` | Master Station Names - station names and (rough) locations | Yes, for names and fallback locations |
| `.FLF` | Fixed Links - interchange/transfer times between stations | Optional (`transfers = TRUE`) |
| `.ALF` | Additional fixed links | Not used |
| `.TSI`, `.SET`, `.DAT`, `.ZTR`, `.REJ` | Ancillary/rejected data | Not used |

Only the `.MCA`, `.MSN` and `.FLF` files are needed to build a GTFS feed.

## Where to get CIF data

There are two sources, and this determines which function you use.

### 1. Rail Delivery Group (via the National Rail Data Portal)

The [National Rail Data Portal](https://opendata.nationalrail.co.uk/) (NRDP)
publishes the Rail Delivery Group's timetable feed - historically known as the
**ATOC** feed. This is the *public passenger* timetable and is the right choice
for most users. It arrives as a zip folder (with a name like `ttis627.zip`)
containing the `.MCA`, `.MSN`, `.FLF` and other files described above.

Convert it with **`atoc2gtfs()`**.

### 2. Network Rail

[Network Rail](https://www.networkrail.co.uk/) publishes its own CIF feed
through its data feeds portal. It contains essentially the same schedules but is
packaged differently - a single gzip-compressed file (for example
`toc-full.CIF.gz`) rather than a zip of separate files - and has small
formatting differences. It also tends to include more operational detail (such
as empty-stock and freight movements).

Convert it with **`nr2gtfs()`**.

> **In short:** RDG/ATOC zip → `atoc2gtfs()`; Network Rail `.CIF.gz` →
> `nr2gtfs()`. The two functions share most of their internals but expect
> different packaging.

## Downloading from the National Rail Data Portal

You need a free account at <https://opendata.nationalrail.co.uk/>. Store your
credentials as `NRDP_username` and `NRDP_password` in your `.Renviron` file -
`usethis::edit_r_environ()` opens it for you. Restart R after editing.

```{r download, eval=FALSE}
library(UK2GTFS)
nrdp_timetable("myfolder/timetable.zip")
```

The companion functions `nrdp_fares()` and `nrdp_routing()` download the fares
and routeing feeds respectively. The fares feed can be converted to GTFS fare
tables alongside the timetable - see [Adding fares](#adding-fares) below. The
routeing guide is not used.

## Converting RDG/ATOC data to GTFS

```{r atoc_example, eval=FALSE}
path_in <- "myfolder/timetable.zip"
gtfs <- atoc2gtfs(path_in = path_in, ncores = 3)
```

```{r atoc_example2, eval=FALSE}
gtfs_write(gtfs, folder = "C:/GTFS", name = "gtfs_rail")
```

This writes `C:/GTFS/gtfs_rail.zip`.

Useful arguments to `atoc2gtfs()`:

* `ncores` - cores for parallel processing (leave one free for the OS).
* `locations` - where to get TIPLOC coordinates; see
  [TIPLOC locations](#tiploc-locations). Default `"tiplocs"` uses the package's
  improved dataset.
* `agency` - where to get `agency.txt`; see [Agency](#agency).
* `transfers` - if `TRUE` (default) build `transfers.txt` from the `.FLF` file.
* `working_timetable` - if `TRUE`, use the Working Timetable (WTT) times (used
  operationally) instead of the public times. Default `FALSE`.
* `public_only` - if `TRUE` (default) keep only calls/services intended for
  public pickup and set-down, dropping empty-stock and non-passenger moves.

## Converting Network Rail data to GTFS

The interface is almost identical, but the input is the `.CIF.gz` file:

```{r nr_example, eval=FALSE}
gtfs <- nr2gtfs(path_in = "toc-full.CIF.gz", ncores = 3)
gtfs_write(gtfs, folder = "C:/GTFS", name = "gtfs_rail_nr")
```

`nr2gtfs()` shares the `locations`, `agency`, `working_timetable` and
`public_only` arguments with `atoc2gtfs()`.

## What the conversion does under the hood

Both functions are wrappers around lower-level importers, which you can call
directly if you need finer control:

* `importMCA()` - reads the main `.MCA` timetable file (the core data).
* `importMSN()` - reads the `.MSN` station file (used when `locations = "file"`
  or to fill in missing TIPLOCs).
* `importFLF()` - reads the `.FLF` fixed-links file (used when
  `transfers = TRUE`).

Other files in the bundle are currently ignored.

## TIPLOC locations

The timetable refers to stations, junctions and other points by **TIPLOC** code.
The `.MSN` file does contain coordinates for these, but they are often only
accurate to about a kilometre and are occasionally badly wrong. `UK2GTFS`
therefore ships an improved `tiplocs` dataset with corrected coordinates, used
by default (`locations = "tiplocs"`).

You have alternatives:

* `locations = "file"` - use the (lower quality) coordinates from the `.MSN`
  file instead.
* Pass your own `sf` data frame of points in the same format as `tiplocs`.
* Pass a path to a CSV formatted like a GTFS `stops.txt`.

If you use the improved dataset but the timetable references a TIPLOC not in it,
`atoc2gtfs()` (with `missing_tiplocs = TRUE`, the default) falls back to the
`.MSN` location and warns you. The improved locations are maintained in the
[UK2GTFS-data](https://github.com/itsleeds/UK2GTFS-data) repository - corrections
are welcome via pull request. You can also build the stops manually from the
station file with `station2stops()`.

## Agency

The CIF files do **not** contain the information needed to build GTFS
`agency.txt` (operator names, URLs, timezones). `UK2GTFS` therefore ships an
[example agency file](https://github.com/itsleeds/UK2GTFS-data/blob/main/data/raw/atoc_agency.csv)
which is used by default (`agency = "atoc_agency"`). You can pass your own data
frame of agency information instead. Contributions to the shared file are welcome
via pull request.

## Adding fares

The timetable feed contains no prices. Fares are published separately by the
Rail Delivery Group as the **fares feed** (specification RSPS5045), which you
can download from the same National Rail Data Portal account with
`nrdp_fares("myfolder/fares.zip")`. The feed is a set of fixed-width text
files (`RJFAF756.FFL`, `.LOC`, `.TTY`, ...) listing every advertised
point-to-point fare in Great Britain - the same data ticket machines use.

The simplest route is to pass the fares zip straight to `atoc2gtfs()`:

```{r fares_onestep, eval=FALSE}
gtfs <- atoc2gtfs("myfolder/timetable.zip", ncores = 3,
                  fares = "myfolder/fares.zip",
                  fares_version = 2)
```

Or do it as a separate step, which lets you inspect the fares data and
convert the same feed several ways:

```{r fares_twostep, eval=FALSE}
fares <- atoc_fares_read("myfolder/fares.zip")
fares$ticket_type   # every ticket code, with description, class and type
fares$railcard      # every railcard code

# GTFS v1: the cheapest standard-class single per station pair
gtfs_v1 <- gtfs_add_railfares(gtfs, fares, fares_version = 1)

# GTFS Fares v2: adult, child and 16-25 Railcard singles and returns
gtfs_v2 <- gtfs_add_railfares(gtfs, fares, fares_version = 2,
                              railcards = "YNG")
```

### How rail fares map onto GTFS

Rail fares are set per **flow**: an origin/destination pair, a fare route
("via any permitted", "not via London", ...) and a ticket type. A flow
endpoint is not always a single station - it can be a **station cluster**
(stations that share fares) or a **group station** such as "LONDON
TERMINALS". This maps onto the two GTFS fare models as follows:

| Fares feed concept | GTFS v1 | GTFS Fares v2 |
|---|---|---|
| Station | `stops.zone_id` (CRS code) | member of `areas`/`stop_areas` |
| Cluster / group station | expanded to member stations | one area with many stops |
| Flow (O/D pair) | `fare_rules` origin/destination | `fare_leg_rules` from/to area |
| Ticket type + price | `fare_attributes` (cheapest only) | `fare_products` |
| Adult / child / railcard | *(not representable)* | `rider_categories` |

Which tickets are converted is controlled by arguments:
`fares_ticket_class` (standard/first), `fares_ticket_type`
(single/return/season) or explicit `fares_ticket_codes`;
`fares_rider_categories` and `fares_railcards` choose the passenger types
(v2 only). By default only **walk-up** tickets - the Anytime, Off-Peak and
Super Off-Peak families - are converted (`fares_walkup_only = TRUE`); the
feed also prices trade, carnet and advance-purchase tier tickets, which are
misleadingly cheap in a journey planner. See `?gtfs_add_railfares` for
details.

### Scenario conversion: a specific travel date, time and booking date

GTFS cannot carry the full restriction model of rail ticketing (validity
varies by date, time, route and train). What the converter *can* do is
evaluate those restrictions for you at conversion time, for one concrete
scenario, and output plain GTFS containing exactly the fares available in
that scenario:

```{r fares_scenario, eval=FALSE}
# Fares for a journey departing 08:00 on Monday 3 August,
# with the ticket bought on 1 July
gtfs <- atoc2gtfs("myfolder/timetable.zip", ncores = 3,
                  fares = "myfolder/fares.zip",
                  fares_version = 2,
                  fares_travel_date = as.Date("2026-08-03"),
                  fares_travel_time = "08:00",
                  fares_booking_date = as.Date("2026-07-01"))
```

Three things happen:

* records are read as valid on the **travel date** (the feed contains
  current and future fares rounds);
* the restriction data (RST file) is evaluated for that date and departure
  time: Off-Peak and Super Off-Peak fares are dropped where their time
  restriction makes them invalid (e.g. departing before 09:30 on a
  weekday), as are fares date-banned on the travel date;
* with a **booking date**, Advance tickets whose booking horizon (TAP
  file) is satisfied are added at their tier prices, instead of being
  excluded.

This is the recommended way to do fare-sensitive analysis (e.g. comparing
peak vs off-peak accessibility, or walk-up vs advance pricing): build one
feed per scenario and compare. Omit `fares_travel_time` to apply only
date-level restrictions; omit `fares_booking_date` for walk-up fares only.

Two caveats. Advance *availability* is quota-controlled per train by the
reservation system and is in no public feed, so Advance prices are the
bookable tiers - a best case. And restriction evaluation covers date bands
and departure-time bands (network-wide or origin-station-specific);
arrival-based, via-based and train-specific restrictions, easements and
minimum-fare windows are treated as "fare remains valid", so the filter
errs towards keeping a fare.

### Limitations and trade-offs

Like the bus fares conversion (see the *Adding Fares - NeTEx to GTFS*
vignette), converting rail fares to GTFS is lossy. Choices are surfaced as
arguments; the losses you cannot buy back are listed here.

1. **GTFS v1 cannot express ticket choice or passenger types.** With
   `fares_version = 1` every station pair gets exactly one price: the
   *cheapest* fare among the ticket types you selected (default:
   standard-class singles). Child, railcard and return fares are ignored.
   Use v2 if you need them.
2. **Ticket time restrictions are not carried into the output.** An
   Off-Peak Single is emitted with its price, but GTFS cannot say it is
   invalid at 08:00, so a journey planner sees an Anytime and an Off-Peak
   product between the same stations and cannot choose correctly. Either
   convert a scenario (`fares_travel_date`/`fares_travel_time` above), or
   restrict conversion to Anytime products (e.g.
   `fares_ticket_codes = c("SDS", "SOS")`).
3. **Returns are priced for the round trip.** GTFS has no native return
   ticket; a return product's `amount` is the price of the whole return
   journey attached to a single leg. Season tickets are excluded by default
   for the same reason (their feed price is the weekly rate).
4. **Advance fares appear only in booking scenarios.** Advance tickets are
   quota-controlled: the flow file lists their price *tiers* with no
   information about which tier is on sale for which train. By default they
   are dropped (`fares_walkup_only = TRUE`); with a
   `fares_booking_date` scenario the bookable tiers are included as
   best-case prices.
5. **Child and railcard prices are calculated, not quoted.** The feed
   derives them from the adult fare via the status discount file
   (typically 50% for children). UK2GTFS applies the percentage and the
   flat-fare/minimum-fare caps, but not the industry rounding rules or
   railcard minimum-fare rules, so a calculated price can differ from the
   retail price by a few pence.
6. **Some flows cannot be placed.** Flow endpoints that are not stations
   (county codes, travelcard zones without member stations) or stations
   absent from your timetable are dropped, with a message reporting the
   count.
7. **TOC-specific ticketing is simplified.** Fares are attached to one
   national "rail" network (v2) or to no route at all (v1), i.e. they are
   treated as inter-available. Operator-only tickets (the TSP file) and
   cross-London rules are not modelled.
8. **Feed size.** The national flow file holds ~700,000 flows and ~7
   million prices. A national v1 conversion expands clusters to station
   pairs and produces a multi-million-row `fare_rules.txt`; v2 stays close
   to the source size because clusters become areas. Both are written fine
   by `gtfs_write()`, but consumers vary in what they accept.

## Shapes

Route shapes (the geographic path a train follows) are not produced by default -
`shapes = FALSE`. Generating rail shapes requires matching schedules to the
physical rail network, which is handled separately by the experimental
`ATOC_shapes()` function.

## Next steps

After conversion, use the tools in the [Working with GTFS files](GTFS.html)
vignette - in particular `gtfs_clean()` and `gtfs_fast_trips()`, which is handy
for spotting schedules with implausible speeds caused by a mislocated TIPLOC.
