---
title: "Adding Fares - NeTEx to GTFS (v1 and v2)"
author: "Malcolm Morgan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding Fares - NeTEx to GTFS (v1 and v2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette explains how to add **fares** to a GTFS feed built with UK2GTFS,
using the NeTEx fare files published by the UK **Bus Open Data Service** (BODS).
It covers where the data comes from, how the functions fit together, and - most
importantly - the conceptual mismatches between NeTEx and GTFS that force some
trade-offs when you convert. The functions described here all live in the
`netex_*` / `gtfs_add_fares*` family and are independent of the existing
timetable conversion code.

(For **heavy rail** fares - the National Rail fares feed used with
`atoc2gtfs()` - see the *Heavy Rail CIF to GTFS* vignette and
`?gtfs_add_railfares`; rail uses a completely different source format.)

## Background: two different worlds

Timetables and fares are published separately in the UK:

* **Timetables** come as **TransXChange** XML and are converted to GTFS with
  [`transxchange2gtfs()`]. This gives you `stops`, `routes`, `trips`,
  `stop_times` and calendars, but **no fares**.
* **Fares** come as **NeTEx** XML (the DfT "fxc" fares profile) in a separate
  BODS *fares* archive. NeTEx is a very large, very general European standard;
  the UK profile uses only a slice of it, but that slice is still far richer
  than anything GTFS can express.

The job of these functions is to read the NeTEx fares, line them up with the
GTFS routes you already built, and write out fares in **either** GTFS fares
specification:

* **GTFS "v1"** - the original `fare_attributes.txt` + `fare_rules.txt` (plus a
  `zone_id` on each stop).
* **GTFS-Fares-v2** - the newer, richer `areas`, `fare_products`,
  `fare_leg_rules`, `rider_categories`, `fare_media` and `networks` tables.

The two are genuinely different data models, so both are supported and you
choose with `fares_version`.

## The NeTEx fare model (what one file contains)

In the BODS fares archive, **one XML file describes the fares for one line, in
one direction, for one product** - for example "Beestons line 91, Inbound,
Adult Single". A single registered service therefore spreads across many files
(adult/child x single/return x inbound/outbound).

The fares themselves are usually a **zone-to-zone fare triangle**:

* the line's stops are grouped into a handful of **fare zones**
  (`FareZone` -> a list of `ScheduledStopPointRef`);
* every **origin-zone / destination-zone pair** points at a **price band**
  (`DistanceMatrixElement` -> `PriceGroupRef`);
* each **price band** has a monetary amount (`PriceGroup` -> `Amount`).

So the meaning is "travelling from zone A to zone B on this line costs GBP X for
this passenger type". A smaller number of files instead describe a single
**flat fare** (one price for the whole line, no zones); these are read as a
single zone-less fare.

## Reading NeTEx

```{r read}
library(UK2GTFS)

# 1. Unpack the BODS fares archive (follows nested zips). Keep the output
#    folder path SHORT on Windows - see "Long file paths" below.
xml_files <- netex_unzip("bodds_fares_archive_20260703.zip", exdir = "C:/nfx")

# 2. Read every file. Reading is CPU-bound, so use several cores.
netex <- netex_read_fares_multiple(xml_files, ncores = 20)

# Files that could not be parsed are reported, not fatal:
netex_read_failures(netex)
```

`netex_read_fares()` reads a single file into a small list of tidy
`data.table`s (`meta`, `stops`, `zones`, `prices`, `fares`) if you want to
inspect one file directly.

## Choosing which fares to convert

A national archive contains many products per line. `netex_fare_types()` shows
you what is available, and `netex_filter_fares()` lets you pick:

```{r choose}
# What products exist?
netex_fare_types(netex)
#>    idx operator_noc line_public_code direction product_name trip_type user_type ...
#> 1:   1         BEES               91   inbound Adult Return    return     adult
#> 2:   2         BEES               91   inbound Adult Single    single     adult
#> 3:   3         BEES               91   inbound Child Single    single     child
#> ...

# Keep only adult single tickets
singles <- netex_filter_fares(netex, trip_type = "single", user_type = "adult")
```

Every attribute shown by `netex_fare_types()` can be used as a filter
(`trip_type`, `user_type`, `product_name`, `direction`, `line_public_code`,
`operator_noc`), and any combination is valid, so all fare types are
selectable.

## Matching fares to routes, and a national report

NeTEx and GTFS do not share route identifiers. The GTFS `route_id` produced by
`transxchange2gtfs()` comes from the TransXChange *ServiceCode*, whereas NeTEx
identifies the line by its public number. Matching is therefore done on:

* NeTEx `operator_noc`  ==  GTFS `agency_id`   (e.g. `BEES`), **and**
* NeTEx `line_public_code`  ==  GTFS `route_short_name`  (e.g. `91`).

```{r match}
gtfs <- transxchange2gtfs("bodds_timetable_archive.zip", ncores = 20)

# How much of the fares data can we actually attach?
netex_fares_report(netex, gtfs)
#> NeTEx fares report
#>   files:            151691  (failed to parse: 0)
#>   operators:        440
#>   lines:            8875
#>   zonal fare files: 77213
#>   flat  fare files: 74478
#>   lines matched to GTFS route: 5204 / 9116
```

The match rate depends entirely on the overlap between the fares archive
(national, 440 operators) and the timetable feed you built. Matching the full
national fares against a regional GTFS naturally leaves most lines unmatched;
against a national timetable feed the rate is much higher.

`netex_fares_report()` is the tool for national runs: it summarises the product
mix and, crucially, how many lines could be matched to a route. Unmatched lines
are returned in `$unmatched`.

### What the national archive looks like

Reading the whole GB Bus Open Data Service fares archive (July 2026) gives a
sense of the scale and variety the functions have to cope with:

| Measure | Value |
|---|---|
| NeTEx fare files | ~151,700 |
| Failed to parse | 0 |
| Operators / lines | 440 / 8,875 |
| Zonal fare files (fare triangle) | ~77,200 |
| Flat fare files (single price) | ~74,500 |
| Flat fares with no line number (operator-wide) | ~42,100 |

Two things stand out. First, roughly **half of all fare files are flat fares**,
so flat-fare support is not an edge case. Second, a large share of those flat
fares carry **no line number** because they apply across an operator's whole
network; these cannot be matched to a single GTFS route by line number and are
reported as unmatched (see limitation 8).

## Producing GTFS fares

The one-step wrapper unpacks, reads, reports and attaches fares:

```{r fromarchive}
# GTFS v1, adult single tickets only
gtfs_v1 <- netex_fares_from_archive(
  "bodds_fares_archive_20260703.zip",
  gtfs          = gtfs,
  fares_version = 1,
  ncores        = 20,
  trip_type     = "single",
  user_type     = "adult"
)
gtfs_write(gtfs_v1, "output", name = "gtfs_with_fares_v1")

# GTFS-Fares-v2, all products at once
gtfs_v2 <- netex_fares_from_archive(
  "bodds_fares_archive_20260703.zip",
  gtfs          = gtfs,
  fares_version = 2,
  ncores        = 20
)
gtfs_write(gtfs_v2, "output", name = "gtfs_with_fares_v2")
```

Or, if you already have the parsed `netex` list, call `gtfs_add_fares()`
(or the lower-level `gtfs_add_fares_v1()` / `gtfs_add_fares_v2()`) directly.

## How NeTEx maps onto GTFS

| NeTEx concept              | GTFS v1                    | GTFS-Fares-v2                         |
|----------------------------|----------------------------|---------------------------------------|
| Fare zone -> stops         | `stops.zone_id`            | `areas` + `stop_areas`                |
| O/D zone pair              | `fare_rules` origin/dest   | `fare_leg_rules` from/to area         |
| Price band amount          | `fare_attributes.price`    | `fare_products.amount`                |
| Line                       | `fare_rules.route_id`      | `networks` + `route_networks`         |
| Passenger type (adult...)  | *(not representable)*      | `rider_categories`                    |
| Ticket medium / channel    | `payment_method` (approx.) | `fare_media`                          |

## Limitations and trade-offs

Converting NeTEx to GTFS is lossy in both directions of the specification.
These are the issues to be aware of; where a choice exists it is exposed as a
function argument.

### 1. GTFS v1 has no passenger categories

`fare_attributes`/`fare_rules` cannot say "this price is for children". GTFS v1
can therefore only carry **one product** cleanly. Select it explicitly, e.g.
`trip_type = "single", user_type = "adult"`. If you leave several products in,
they will all be emitted and a rider-facing planner will simply apply the
cheapest matching fare, which is usually wrong. **Use GTFS-Fares-v2 if you need
adult *and* child, or single *and* return.**

### 2. GTFS v1 allows only one fare zone per stop

`stops.zone_id` is a single value. But a stop can legitimately sit in different
fare zones on different lines, in different directions, or (in the source data)
even within one file. When conversions collide, `gtfs_add_fares_v1()` keeps the
first assignment, drops the now-orphaned fare rules, and warns. At national
scale this is severe - converting the whole archive at once reports on the order
of 80,000 conflicted stops - which is exactly why v1 is best used per route or
per direction. The practical advice is to build v1 feeds **one direction at a
time** (`direction = "inbound"` or `"outbound"`), or to use v2, where
`stop_areas` is many-to-many and a stop can belong to as many areas as needed.

### 3. Directional fares

NeTEx fares are directional (the triangle is not symmetric). GTFS networks are
not direction-aware, so both directions' leg rules attach to the same network.
For typical symmetric bus fares this is harmless; where inbound and outbound
prices genuinely differ, a planner cannot always pick the right one. Areas are
namespaced by direction so the data is at least internally consistent.

### 4. Only stops present in the GTFS feed get fares

A fare zone may list stops that are not in the matched GTFS route (the route
version you converted may serve fewer stops). Zones with no stop in the feed,
and any fare rule that references them, are dropped so the output has no
dangling references.

### 5. Flat fares vs zonal fares

About half of the national archive uses **flat fares** (one price for the whole
line) rather than a zonal fare triangle. These are read (`meta$fare_kind` is
`"flat"`) and converted to a single route-wide fare: in v1 a `fare_rule` with a
`route_id` but no origin/destination; in v2 a `fare_leg_rule` on the network
with no `from_area_id`/`to_area_id`. No `zone_id`/`areas` are produced for a
flat-fare line.

### 6. Operator-wide fares cannot be matched by line

Many flat fares (about 42,000 files nationally) carry **no line number** because
they apply to a whole operator's network, not a single service. Matching is by
operator + line number, so these have nothing to match against and appear in the
report's `$unmatched`. Attaching them would require a policy decision (apply to
*every* route of the operator), which is deliberately left to the caller rather
than assumed.

### 7. One product/tariff per file is read

Each NeTEx file is assumed to carry one line/direction/product with a single
tariff, which matches the BODS export convention. Files that pack multiple
tariffs are read using the first.

### 8. Payment method and media are approximated

The rich NeTEx `SalesOfferPackage` (distribution channel, payment methods,
ticket type) is collapsed to a GTFS `payment_method` (v1) or a single
`fare_media` row (v2). Both default to on-board cash payment and are exposed as
arguments (`payment_method`, `fare_media_name`, `fare_media_type`).

### 9. Long file paths on Windows

BODS NeTEx file names routinely exceed 100 characters. Nested inside long
operator-folder names they blow past the Windows 260-character path limit,
which makes extraction silently fail. `netex_unzip()` avoids this by extracting
nested archives into short, sequentially-named folders - but keep your `exdir`
short as well (a short drive-root path such as `C:/nfx` is safest).

## Parallel processing and scale

Reading the ~150,000-file national NeTEx archive is by far the expensive step
and is parallelised with `furrr`/`future`. Pass `ncores` to
`netex_read_fares_multiple()` (or the `netex_fares_from_archive()` wrapper) to
spread the work over up to ~20 cores, with a progress bar. On a 20-core machine
the full national read takes on the order of an hour; a single operator is a
few seconds.

The matching step (`netex_match_routes()`) is vectorised and handles tens of
thousands of routes without a per-file loop. The conversion step
(`gtfs_add_fares_*()`) runs single-core but is fast, and it silently skips files
whose line does not match any route (reporting the count once) so that running a
national fares archive against a regional GTFS does not flood the log.

`transxchange2gtfs()` has its own `ncores` argument for building the timetable
feed in parallel.
