---
title: "Working with GTFS files - Cleaning, Manipulating and Analysing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with GTFS files - Cleaning, Manipulating and Analysing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

As well as converting UK formats to GTFS, `UK2GTFS` provides a toolkit for
reading, cleaning, checking, reshaping and analysing GTFS feeds - whether they
came from `UK2GTFS` itself or from anywhere else.

## Background: what is GTFS, and what is a "gtfs object"?

The [General Transit Feed Specification](https://gtfs.org/) (GTFS) is a `.zip`
file containing a set of related CSV tables - `agency.txt`, `stops.txt`,
`routes.txt`, `trips.txt`, `stop_times.txt`, `calendar.txt`,
`calendar_dates.txt`, and optional extras such as `shapes.txt` and
`frequencies.txt`. The tables are linked by ID columns (a trip belongs to a
route, has many stop_times, and runs on a service_id calendar).

In `UK2GTFS` a GTFS feed is held in memory as a **named list of data frames** -
one element per table, named after the file (`gtfs$stops`, `gtfs$trips`, and so
on). Every function below takes and/or returns such an object, so you can chain
them together and inspect any table directly as an ordinary data frame.

```{r read, eval=FALSE}
library(UK2GTFS)

gtfs <- gtfs_read("myfeed.zip")
names(gtfs)      # the tables present
head(gtfs$stops) # inspect any table as a data frame
```

## Reading and writing

| Function | Purpose |
|----------|---------|
| `gtfs_read(path)` | Read a GTFS `.zip` into a gtfs object |
| `gtfs_write(gtfs, folder, name)` | Write a gtfs object back to a `.zip` |

```{r readwrite, eval=FALSE}
gtfs <- gtfs_read("myfeed.zip")
# ... work on it ...
gtfs_write(gtfs, folder = "C:/GTFS", name = "myfeed_clean")
```

`gtfs_write()` has options to strip commas, tabs and newlines from text fields
(`stripComma`, `stripTab`, `stripNewline`) - useful because stray delimiters in
operator or stop names can break strict GTFS consumers.

## Cleaning and validating

Real-world feeds - and conversions of messy source data - contain errors. Two
families of function help.

### `gtfs_clean()` - fix common errors

`gtfs_clean(gtfs)` removes a set of well-known problems, including: stops with no
coordinates or no location information; trips with fewer than two stops; stops
that are never used; and (with `public_only = TRUE`) non-public services such as
freight or empty-stock moves that have no GTFS route type and would otherwise
break tools like OpenTripPlanner. It also tidies up orphaned shapes and
frequencies.

```{r clean, eval=FALSE}
gtfs <- gtfs_clean(gtfs)
```

### `gtfs_validate_internal()` - check for problems

`gtfs_validate_internal(gtfs)` checks the feed against the GTFS specification
and reports what it finds at three severities: **Error** (breaks the spec),
**Warning** (probably a mistake) and **Note** (worth knowing). Checks cover
every table - including shapes, frequencies, transfers and the fare tables -
and include: required tables and columns, duplicated ids, referential
integrity of every cross-table reference (for example stop_times referencing
a stop that isn't in `stops.txt`), coordinate ranges, enum and date/time
values, times that go backwards along a trip, and services that can never
run. It reports problems but does not change the data; it invisibly returns
a data frame of the problems for programmatic use.

```{r validate, eval=FALSE}
gtfs_validate_internal(gtfs)
problems <- gtfs_validate_internal(gtfs)  # severity / table / message
```

### `gtfs_force_valid()` - make it valid by removing problems

Where `gtfs_validate_internal()` only reports, `gtfs_force_valid(gtfs)` acts: it
drops stops with missing locations, routes not present in `agency`, trips not in
`routes`, stop_times not in `trips` or `stops`, and calendar entries with no
matching trips. Use it as a last resort to guarantee referential integrity when
a downstream tool refuses a slightly broken feed.

### `gtfs_fast_trips()` and `gtfs_fast_stops()` - catch mislocated stops

A frequent, sneaky error is a stop placed in the wrong location, which makes a
vehicle appear to "teleport" across the country between two calls.
`gtfs_fast_trips(gtfs)` returns the `trip_id`s whose fastest segment exceeds
`maxspeed` (default 83 m/s ≈ 185 mph, the top speed on HS1); `gtfs_fast_stops()`
returns an `sf` data frame of the suspect stops for mapping.

```{r fast, eval=FALSE}
bad_trips <- gtfs_fast_trips(gtfs, maxspeed = 45) # ~100 mph, for a bus feed
bad_stops <- gtfs_fast_stops(gtfs)
```

## Manipulating

### Merging

`gtfs_merge(gtfs_list)` combines a list of gtfs objects into one, handling ID
clashes (if duplicate IDs are found across feeds it regenerates fresh IDs). If
small inconsistencies (such as an operator name spelt two ways) block a clean
merge, pass `force = TRUE` to take the first instance as authoritative.

```{r merge, eval=FALSE}
combined <- gtfs_merge(list(gtfs_bus, gtfs_rail), force = TRUE)
```

### Splitting

`gtfs_split(gtfs, n_split = 2)` divides a large feed into a list of smaller,
size-balanced feeds (split by `agency_id`), which is helpful when a feed is too
big for a downstream tool. `gtfs_split_ids(gtfs, trip_ids)` splits on a specific
set of `trip_id`s, returning a `$true` feed (matching trips) and a `$false` feed
(everything else).

```{r split, eval=FALSE}
parts <- gtfs_split(gtfs, n_split = 4)
sel   <- gtfs_split_ids(gtfs, trip_ids = bad_trips)
```

### Clipping to an area

`gtfs_clip(gtfs, bounds)` restricts a feed to a geographic area. `bounds` is an
`sf` polygon (or multi-polygon) in CRS 4326. Trips crossing the boundary are
truncated at the edge, and trips that call only once inside the area are dropped.
The optional tables - shapes, frequencies, transfers, pathways and the fare
tables - are pruned so they only reference the stops, routes and trips that
remain.

```{r clip, eval=FALSE}
library(sf)
leeds <- st_read("leeds_boundary.geojson") # must be EPSG:4326
gtfs_leeds <- gtfs_clip(gtfs, bounds = leeds)
```

### Trimming to a date range

`gtfs_trim_dates(gtfs, startdate, enddate)` keeps only services that run between
two dates and shrinks the calendars accordingly - useful for cutting a
multi-year feed down to the week or month you care about. Frequencies and
shapes belonging to removed trips are removed with them.

```{r trim, eval=FALSE}
gtfs_march <- gtfs_trim_dates(gtfs,
                              startdate = as.Date("2024-03-01"),
                              enddate   = as.Date("2024-03-31"))
```

### Compressing

`gtfs_compress(gtfs)` reduces file size by replacing the long, human-readable IDs
that `UK2GTFS` preserves during conversion (such as NaPTAN stop codes) with
compact integers. Every table that references a rewritten id - stop_times,
transfers, pathways, frequencies, shapes, stop_areas, fare_rules and
route_networks - is updated together, so the feed stays consistent. Do this
once you no longer need to trace IDs back to the source data.

```{r compress, eval=FALSE}
gtfs <- gtfs_compress(gtfs)
```

## Analysing

### Quick summary

`gtfs_summary(gtfs)` prints the number of rows in each table and the date range
covered by the calendar - the fastest way to sanity-check a feed.

```{r summary, eval=FALSE}
gtfs_summary(gtfs)
```

### Service frequency at each stop

`gtfs_stop_frequency(gtfs, startdate, enddate)` counts how many trips call at
each stop over a date range, returning the `stops` table with `stops_total` and
`stops_per_week` columns added. This accounts for the calendar *and* calendar
exceptions, so it reflects services actually running - a good proxy for "how well
served is this stop?".

```{r freq, eval=FALSE}
stops <- gtfs_stop_frequency(gtfs,
                             startdate = as.Date("2024-03-01"),
                             enddate   = as.Date("2024-03-31"))
```

> Note the date range must fall inside the feed's calendar or you will get no
> results - a common gotcha with short feeds such as those from the
> [NPTDR](NPTDR.html).

### Trips per zone, by time of day and mode

`gtfs_trips_per_zone(gtfs, zone, ...)` aggregates trips into your own set of
geographic zones (an `sf` polygon data frame), broken down by day of week, time
band and (optionally) mode. The default time bands split the day into Night,
Morning Peak, Midday, Afternoon Peak and Evening, and you can supply your own via
the `time_bands` argument.

```{r zone, eval=FALSE}
library(sf)
zones <- st_read("msoa_zones.geojson")

trips <- gtfs_trips_per_zone(
  gtfs,
  zone      = zones,
  zone_id   = "MSOA_code",
  startdate = as.Date("2024-03-04"),
  enddate   = as.Date("2024-03-10"),
  by_mode   = TRUE
)
```

## Mapping

Three helpers turn GTFS tables into `sf` objects you can plot with `plot()`,
`ggplot2` or `tmap`:

| Function | Returns |
|----------|---------|
| `gtfs_stops_sf(gtfs)` | Stops as points |
| `gtfs_trips_sf(gtfs)` | Every trip as a line |
| `gtfs_routes_sf(gtfs)` | One representative line per route (lighter to plot) |

```{r sf, eval=FALSE}
library(sf)
stops_sf  <- gtfs_stops_sf(gtfs)
routes_sf <- gtfs_routes_sf(gtfs)
plot(st_geometry(routes_sf))
```

## Fixing uneven stop times

Some bus timetables give several consecutive stops the *same* time rather than a
unique time each. `gtfs_interpolate_times(gtfs)` spreads these out, assigning each
stop a distinct arrival and departure time by interpolating between the known
times - which makes speed and journey-time analysis far more meaningful.

```{r interp, eval=FALSE}
gtfs <- gtfs_interpolate_times(gtfs)
```

## A typical workflow

Putting it together, a common end-to-end pattern is:

```{r workflow, eval=FALSE}
library(UK2GTFS)

gtfs <- gtfs_read("myfeed.zip")   # 1. read
gtfs_summary(gtfs)                #    look
gtfs <- gtfs_clean(gtfs)          # 2. clean
gtfs_validate_internal(gtfs)      # 3. check what remains

gtfs <- gtfs_clip(gtfs, bounds = my_area)              # 4. subset to area
gtfs <- gtfs_trim_dates(gtfs, start_date, end_date)    #    and to dates

stops <- gtfs_stop_frequency(gtfs, start_date, end_date) # 5. analyse

gtfs_write(gtfs, folder = "C:/GTFS", name = "myfeed_clean") # 6. save
```

## See also

* [TransXChange to GTFS](transxchange.html) - converting bus, coach, tram, metro
  and ferry timetables.
* [Heavy Rail CIF to GTFS](ATOC.html) - converting train timetables.
* [NPTDR to GTFS](NPTDR.html) - converting historic all-mode archives.
