---
title: "NPTDR to GTFS - Historic Timetables"
author: "Malcolm Morgan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NPTDR to GTFS - Historic Timetables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explains how to convert the **National Public Transport Data
Repository** (NPTDR) archives into [GTFS](https://gtfs.org/). NPTDR is the only
practical source of a *single, whole-of-Britain, all-modes* historic snapshot,
which makes it valuable - and slightly awkward - to work with.

## Background: what is the NPTDR?

The National Public Transport Data Repository was an annual snapshot of **every**
public transport journey in Great Britain - bus, coach, rail, tram, metro and
ferry - taken during a selected week each October. It was produced between
roughly 2004 and 2011, after which it was discontinued (its role was effectively
taken over by the live Traveline and, later, Bus Open Data feeds).

Its distinguishing features:

* **Comprehensive and multi-modal.** One archive covers the whole of Great
  Britain across all modes in a consistent format - unusual, and the main reason
  to use it.
* **Historic only.** The newest snapshot is from around 2011, so NPTDR is for
  studying the past (network change over time, services that no longer exist,
  before/after analyses), not for current journey planning.
* **A snapshot of one week.** Each archive represents a single sample week, so
  the calendar it produces is short.

### Why the format is fiddly

NPTDR distributes the timetable in the **CIF** format - but a *different dialect*
from the heavy-rail CIF covered in the [Heavy Rail CIF](ATOC.html) vignette.
This NPTDR/ATCO-CIF dialect uses record types beginning with `Q` (`QS` service
header, `QO` origin, `QI` intermediate, `QT` terminating, `QL`/`QB` locations,
`QE` exceptions, `QR` repetitions). An NPTDR archive is a nested set of zips:

* one CIF file **per administrative area per mode**, grouped in `Admin_Area`
  zips (so a single archive contains many CIF files);
* a **NaPTAN** stops export for stop locations;
* various supporting files.

`UK2GTFS` unpacks all of this for you.

## Where to get the data

The historic NPTDR archives are no longer served from a single official portal.
Copies are held at the [UK Data Archive](https://www.data.gov.uk/) and various
mirrors; you download a per-year zip (for example `October-2006.zip`). A small
sample archive suitable for testing the package can be downloaded with:

```{r dl_example, eval=FALSE}
library(UK2GTFS)
dl_example_file(".", type = "nptdr")
```

## Converting NPTDR to GTFS

The single entry point is `nptdr2gtfs()`:

```{r nptdr_example, eval=FALSE}
library(UK2GTFS)

gtfs <- nptdr2gtfs(path = "October-2006.zip")

gtfs_write(gtfs, folder = "C:/GTFS", name = "gtfs_nptdr_2006")
```

Arguments:

* `path` - path to the NPTDR year zip.
* `enhance_stops` - if `TRUE` (default) the **current** NaPTAN is downloaded and
  used to fill in any stops that are missing locations in the historic archive.
  See the note below.
* `naptan` - the NaPTAN stop locations, by default fetched with `get_naptan()`.
  Pass a pre-loaded copy to avoid re-downloading.
* `n_files` - a debugging aid: an integer vector selecting a subset of the CIF
  files (for example `1:10`) so you can test on a fraction of the archive before
  running the whole thing.
* `silent` - suppress progress messages.

Because an archive contains many CIF files covering the whole country, a full
conversion processes a lot of data - test with `n_files` first.

## Stop locations, old and new

The archive ships its own NaPTAN export, and the CIF files themselves carry
approximate coordinates (in British National Grid easting/northing, which
`UK2GTFS` reprojects to WGS84). These historic locations are used first.

Some stops referenced in the timetable will still lack a location - typically
old, temporary or moveable stops. With `enhance_stops = TRUE` the **present-day**
NaPTAN is consulted to fill gaps. Be aware of the trade-off: this improves
completeness but mixes a modern stop location into a historic dataset, and stops
that have since been removed simply cannot be recovered. Any that remain
unmatched trigger a warning; you can strip them out afterwards with
`gtfs_clean()`:

```{r nptdr_clean, eval=FALSE}
gtfs <- gtfs_clean(gtfs)
```

## Things to watch for

* **Short calendars.** Because each archive samples a single week, the resulting
  GTFS `calendar`/`calendar_dates` cover a narrow date range. When you analyse
  the feed (see [Working with GTFS files](GTFS.html)) make sure your analysis
  dates fall inside that week, or you will get empty results.
* **Historic bank holidays.** The conversion uses a bundled
  `historic_bank_holidays` dataset to resolve operating profiles correctly for
  the year in question.
* **Data quality.** These are old files with occasional malformed times and
  missing stops; the importer repairs the common cases and warns about the rest.
  Expect to run `gtfs_clean()` and `gtfs_validate_internal()` afterwards.

## Next steps

Historic feeds are most useful for analysis and comparison. See the [Working
with GTFS files](GTFS.html) vignette for counting services
(`gtfs_stop_frequency()`, `gtfs_trips_per_zone()`), clipping to an area
(`gtfs_clip()`) and validating the result.
