---
title: "TransXchange - Bus, Tram, and Tube Timetables"
author: "Malcolm Morgan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TransXchange}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vigneete outlines how to produce GTFS timetables for busses and light rail network using TransXchange files.

## TransXchange Data - for bus, tram, and metro

Data for other modes can be downloaded from https://www.travelinedata.org.uk/traveline-open-data/data-reporting/ you will need to apply for access to the FTP sever to download the data.

The data comes as zipped folders for different regions of the UK.

## Converting TransXchange to GTFS

To do a simple conversion to GTFS, load the package, and use the `transxchange2gtfs()` function. The function requires the following arguments.

* `path_in` is the file path to either a single zip folder or many XML files
* `path_out` is the path to a folder where the results will be saved
* `name` is the file name of the result by default `gtfs.zip`
* `ncores` If > 1 then multicore processing is used to speed up the processing. Alwasy leave one spare core for your OS.

A simple example is shown below:

```{r setup, eval =  FALSE}
library(UK2GTFS)
path_in <- "E:/OneDrive - University of Leeds/Data/UK2GTFS/TransXchange/timetable"
path_in <-  file.path(path_in,"EA.zip")
path_out = "E:/GTFS"
transxchange2gtfs(path_in = path_in,
                  path_out = path_out,
                  name = "EA_gtfs",
                  ncores = 4)

```


```{r test, eval =  FALSE, echo=FALSE}
# Setup
#onedrive <- "C:/Users/malco/OneDrive - University of Leeds/"
onedrive <- "E:/OneDrive - University of Leeds/"
ncores <- 5


library(UK2GTFS)
path_in <- paste0(onedrive,"Data/UK2GTFS/TransXChange/data_20201102/")
files <- list.files(path_in, full.names = FALSE, pattern = "zip")
files <- files[files != "NCSD.zip"]

cal = get_bank_holidays()
naptan = get_naptan()

for(i in 1:length(files)){
  message(files[i])
  file_in <-  file.path(path_in,files[i])
  path_out <- paste0(onedrive,"Data/UK2GTFS/GTFS/gtfs_20201102")
  name_out <- gsub(".zip","",files[i], fixed = TRUE)
  gtfs <- transxchange2gtfs(path_in = file_in,
                  path_out = path_out,
                  name = name_out,,
                  cal = cal,
                  naptan = naptan,
                  ncores = ncores)
}



```
