---
title: "TransXChange to GTFS - Bus, Coach, Tram, Metro and Ferry"
author: "Malcolm Morgan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TransXChange to GTFS - Bus, Coach, Tram, Metro and Ferry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explains how to convert **TransXChange** timetables into
[GTFS](https://gtfs.org/), the difference between the two main sources of
TransXChange data (Traveline and the Bus Open Data Service), and the pitfalls to
watch out for.

## Background: what is TransXChange?

TransXChange (often abbreviated *TXC*) is the UK Department for Transport's XML
standard for describing bus, coach, tram, metro and ferry timetables. It is the
format that operators must use to register services with the Traffic
Commissioners, so almost every scheduled road-based service in Great Britain is
described in TransXChange at some point in its life.

Because it was designed for *registration* rather than for *journey planning* it
has some awkward properties:

* It is verbose XML. A single medium-sized operator can run to tens of thousands
  of files.
* One file describes one *service*, and the same real-world journey is often
  spread across many files (and re-uploaded every time the timetable changes -
  see [Duplicate and superseded files](#duplicate-and-superseded-files) below).
* **It does not contain stop coordinates.** Stops are referenced only by their
  ATCO code, and you must look the location up in the separate NaPTAN database
  (see [Stop locations - the NaPTAN](#stop-locations---the-naptan)).
* Operating patterns are described with rules ("every weekday except bank
  holidays") rather than explicit dates, so a calendar has to be computed.

GTFS, by contrast, is a flat set of CSV files designed to be consumed directly
by journey planners such as [OpenTripPlanner](https://www.opentripplanner.org/),
[R5](https://github.com/conveyal/r5) and Google Maps. Converting TransXChange to
GTFS makes the data usable by that wide ecosystem of tools.

## Where to get TransXChange data

There are two national sources. They contain broadly the same underlying
registrations but differ in packaging, freshness and coverage.

### 1. Traveline National Dataset (TNDS)

[Traveline](https://www.travelinedata.org.uk/traveline-open-data/data-reporting/)
publishes the *Traveline National Dataset* (TNDS). You must apply for free
access to their FTP server. The data arrives as zipped folders, one per region
of Great Britain (for example `EA.zip` for East Anglia, `S.zip` for Scotland).

* **Pros:** organised cleanly by region; includes coach, some ferry and light
  rail; a long-established, relatively stable feed.
* **Cons:** each region is a separate download; updated less frequently than
  BODS; being progressively superseded by BODS as the DfT's preferred source.

### 2. Bus Open Data Service (BODS)

The [Bus Open Data Service](https://data.bus-data.dft.gov.uk/) (BODS) is the
DfT's newer, statutory feed. Since 2020 bus operators in England are legally
required to publish their timetables here, so it is the most complete and
up-to-date source for English buses. You can download TransXChange directly, or
download a ready-made national **GTFS** file produced by
[ITO World's](https://www.itoworld.com) own converter.

* **Pros:** most current data; legally mandated so very complete for England;
  offers a ready-made GTFS download.
* **Cons:** English buses only (Scotland and Wales still rely on TNDS for full
  coverage); the bulk "change" archives accumulate **many superseded versions of
  the same service**, which will inflate trip counts if converted naively (see
  below); tram, metro and ferry are patchy.

### Which should I use?

* If you just want current **English bus** GTFS and are not fussy about the
  details of conversion, download the ready-made GTFS from BODS - it is the
  least effort.
* If you need **tram, metro, ferry or coach**, need **Scotland or Wales**, want
  a second independent conversion for comparison, or need **historical**
  timetables, use `UK2GTFS`. The package adds extra data checks and supports the
  non-bus modes that the ready-made BODS GTFS omits.

## Converting TransXChange to GTFS

The workhorse is `transxchange2gtfs()`. In its simplest form:

```{r trans_example, eval=FALSE}
library(UK2GTFS)

path_in <- "EA.zip" # a TNDS region zip, or a folder of .xml files
gtfs <- transxchange2gtfs(path_in = path_in, ncores = 3)
```

Key arguments:

* `path_in` - a single zip folder, or a character vector of paths to `.xml`
  files.
* `ncores` - if `> 1`, multi-core processing is used to speed things up. Always
  leave one core free for the operating system.
* `try_mode` - if `TRUE` (the default) files that fail to convert are skipped
  rather than aborting the whole run. This is robust, but be aware the result
  may be missing some routes - check the messages.
* `cal` / `naptan` - the bank-holiday calendar and stop locations. By default
  these are fetched for you with `get_bank_holidays()` and `get_naptan()`. If
  you are converting many files it is much faster to fetch them once and pass
  them in (see below).
* `scotland` - `"auto"`, `"yes"` or `"no"`. Controls whether Scottish bank
  holidays are used. `"auto"` treats a file ending in `S.zip` as Scottish.
* `filter_duplicate_files` - see [Duplicate and superseded
  files](#duplicate-and-superseded-files).

Once converted, save the GTFS to disk:

```{r trans_example2, eval=FALSE}
gtfs_write(gtfs, folder = "C:/GTFS", name = "gtfs_EA")
```

This writes `C:/GTFS/gtfs_EA.zip`.

### Converting many regions efficiently

`get_naptan()` and `get_bank_holidays()` download data from the internet. When
looping over several regional zips, fetch them once and reuse them:

```{r trans_loop, eval=FALSE}
library(UK2GTFS)

naptan <- get_naptan()
cal    <- get_bank_holidays()

files <- list.files("path/to/tnds", pattern = "\\.zip$", full.names = TRUE)

for (f in files) {
  message(f)
  gtfs <- transxchange2gtfs(
    path_in = f,
    cal     = cal,
    naptan  = naptan,
    ncores  = 3
  )
  name_out <- gsub(".zip", "", basename(f), fixed = TRUE)
  gtfs_write(gtfs, folder = "path/to/output", name = name_out)
}
```

## How the conversion works

`transxchange2gtfs()` is a convenience wrapper around several lower-level steps,
which you can also call individually:

1. `transxchange_import()` reads each XML file into R data frames.
2. `transxchange_export()` turns those into a GTFS object.
3. `gtfs_merge()` combines the per-file GTFS objects into one.
4. `gtfs_write()` saves the result.

Each TransXChange file is converted into its own small GTFS object and these are
then merged. Understanding this matters because the merge is where problems tend
to surface.

## Stop locations - the NaPTAN

TransXChange files reference stops only by their ATCO code; they do **not**
contain coordinates. `UK2GTFS` therefore looks each stop up in the
[NaPTAN](https://www.gov.uk/government/publications/national-public-transport-access-node-schema)
(National Public Transport Access Nodes), the DfT's database of every public
transport stop in Great Britain.

`get_naptan()` always downloads the latest copy. The NaPTAN is not perfect - it
misses some stops and has wrong locations for others - so `UK2GTFS` ships two
internal correction datasets, `naptan_missing` (stops absent from the NaPTAN)
and `naptan_replace` (better coordinates for known-bad stops), which are applied
automatically. Contributions of new or corrected stops are welcome via the
[UK2GTFS-data](https://github.com/itsleeds/UK2GTFS-data) repository.

## Merging problems

Because each file is converted separately and then merged, the merge can fail on
small inconsistencies between files - most commonly when the same operator is
spelt two different ways (for example `"Yorkshire Coastliner Ltd"` versus
`"Yorkshire Coastliner"`), which looks like two agencies sharing one ID.

If `transxchange2gtfs()` cannot merge cleanly it returns a **list** of GTFS
objects instead of a single object. You can then merge them manually, forcing
past the conflict:

```{r trans_merge, eval=FALSE}
gtfs <- gtfs_merge(gtfs_list, force = TRUE)
```

`force = TRUE` keeps the first instance of each duplicated ID, which is almost
always the correct behaviour. You can also set `force_merge = TRUE` in the
original `transxchange2gtfs()` call to do this automatically.

## Duplicate and superseded files

This is the single most important thing to understand when working with **bulk
archives** (especially the BODS change archives).

Every time an operator revises a timetable they upload a **new** TransXChange
file for the same `ServiceCode`, but the older, superseded files usually remain
in the archive and often still declare an open-ended operating period. If you
convert all of them, the same physical bus journey appears once *per file
version*, so counting trips on a given date over-estimates service - in one test
archive a single route appeared five times.

A normal single download of *current* data does not have this problem (each
service appears once), so you only need to worry about this for accumulating
archives.

Set `filter_duplicate_files = TRUE` to keep only the operative version of each
service:

```{r trans_filter, eval=FALSE}
gtfs <- transxchange2gtfs(
  path_in                = "bods_archive.zip",
  filter_duplicate_files = TRUE,
  filter_date            = as.Date("2024-06-01"), # a date inside the period you care about
  ncores                 = 3
)
```

For historical analysis, set `filter_date` to a date within the period you are
studying - the filter keeps the version that was operative on that date. Under
the hood this calls `txc_filter_files()`, which you can also run directly on a
vector of file paths; see `?txc_filter_files` for the exact rules and their one
documented limitation (trip counts can still double-count *after* the next
scheduled timetable change).

## Comparison with the ready-made BODS GTFS

`UK2GTFS` output has been validated against BODS's own (ITO World) GTFS. Where
both feeds cover the same route, agreement is high - around 80% of shared
operator/line pairs had identical daily trip counts and matching departure
times. The remaining differences are mostly about *which input files are
included* (superseded revisions, out-of-region services, and coach/ferry
datasets that live outside the TransXChange archive) rather than the conversion
itself. In short: both are reasonable, and having an independent open-source
converter is useful precisely because TransXChange-to-GTFS involves
interpretation.

## Next steps

Once you have a GTFS file you will usually want to clean and check it. See the
[Working with GTFS files](GTFS.html) vignette for `gtfs_clean()`,
`gtfs_validate_internal()`, `gtfs_clip()` and the analysis functions.
